Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-02 paper/main.tex
caption, references

\subsection{Metrics}
We use two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with AUGUSTUS and further explained in Section \ref{sec:augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. (See Table \ref{tab:statistics_all} for the class distribution and Tables S1 and S2 for a more in-depth report our model performance including the intergenic class). In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes.

@@

The final predictions of a single model are constructed by overlapping predictions, that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see Figure \ref{fig:length_wise_bias}). It also improves the average model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. The figures in Section 4 of the Supplements show the effect of overlapping for each genome, ordered by N75. We found that overlapping tends to work best if the genomes are not very fragmented and we used it for both animals and plants.

@@

\begin{figure*}[!tpb]
\label{fig:results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results.  {\bf a)} Boxplots comparing the Subgenic F1 scores for all animal and plant genomes that were not used during training of Helixer, which is equivalent to all validation and test genomes. Results are given for Helixer and AUGUSTUS with the median predictions scores being denoted by the red lines. Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf c)} Average Genic F1 and basepair wise accuracy with respect to the position in the 20000 basepair long input sequences of all animals on top and all plant genomes below shown with and without overlapping. Each value is the average performance of a 200 basepair long chunk by one of the final ensemble models. The regular metrics show a significant negative bias towards the borders of each sequence. Overlapping the predictions helps to mitigate a lot of that while showing no negative side effects elsewhere on average. See Section \ref{sec:overlapping} for a breakout of the effect of overlapping by each individual species.}
\end{figure*}

\begin{table}[!t]
\processtable{Experimental Results and Comparison\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
AUGUSTUS & 0.632 & 0.757 \\
Dilated CNN Baseline & -  & - \\
Best Helixer model & -  & -  \\
Best model + varied input length & -  & - \\
Best model (+ varied input length) + overlapping &  0.85  & 0.843  \\
(Varied input length +) overlapping + ensembling & 0.868  & 0.863  \\
\botrule
\end{tabular}}{{\it Note:} Values are the median in Subgenic F1 across all validation and test genomes of the respective group. Varied input length was only used in the animal case. The best model was chosen out of the eight models of the ensemble for having the best performance in the reported setup.}
\end{table}

\section{Results}
Figure \ref{fig:boxplot} shows the side-by-side comparison of the distribution of performances on all non-training genomes in the animal and plant case by Subgenic F1. In the case of Helixer, these scores represent cross-species predictions. We also compare the median performances of different configurations of Helixer with AUGUSTUS and a dilated CNN baseline in Table \ref{tab:results}. The results show a clear improvement over the AUGUSTUS both in median performance and spread. We also significantly outperform a baseline constructed of a dilated CNN architecture.

>>>>

\subsection{Metrics}
We mainly use two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with AUGUSTUS and further explained in Section \ref{sec:augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. (See Table \ref{tab:statistics_all} for the class distribution and Tables S1 and S2 for a more in-depth report our model performance including the intergenic class). In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes.

In Figure \ref{fig:main_results}c we also report on the overall base pair level accuracy besides the Genic F1.

@@

The final predictions of a single model are constructed by overlapping predictions, that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the average model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. The figures in Section 4 of the Supplements show the effect of overlapping for each genome, ordered by N75. We found that overlapping tends to work best if the genomes are not very fragmented and we used it for both animals and plants.

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top row shows the animal data; the bottom row is concerned with the plants. Yellow dots are for AUGUSTUS and the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 scores for all genomes that were not used during training of Helixer, which is equivalent to all validation and test genomes. Each dot represents the prediction performace for one genome. The median prediction scores are shown by the red line. Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} Scatterplots showing the prediction performances measured in Subgenic F1 by genome length. The regression line is shown with a 95\% confidence interval. {\bf c)} The average Genic F1 and basepair wise accuracy with respect to the position in the 20000 basepair long input sequences. The dashed line shows the same without overlapping. Each value is the average performance of a 200 basepair long subsection by one of the models of the final ensemble. The usage of the model without overlapping results in a significant negative bias towards the borders of each sequence. Overlapping the predictions helps to mitigate a lot of that while showing no negative side effects elsewhere on average. See Section 4 for the Supplements for the effects of overlapping for each individual species.}
\end{figure*}

\begin{table}[!t]
\processtable{Summary of Experimental Results\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
AUGUSTUS & 0.632 & 0.757 \\
Dilated CNN Baseline & -  & - \\
Best Helixer model & -  & -  \\
Best model + varied input length & -  & - \\
Best model (+ varied input length) + overlapping &  0.85  & 0.843  \\
(Varied input length +) overlapping + ensembling & 0.868  & 0.863  \\
\botrule
\end{tabular}}{{\it Note:} Values are the median in Subgenic F1 across all validation and test genomes of the respective group. Varied input length was only used in the animal case. The best model was chosen out of the eight models of the ensemble for having the best performance in the reported setup.}
\end{table}

\section{Results}
Figure \ref{fig:main_results}a shows the side-by-side comparison of the distribution of performances on all non-training genomes in the animal and plant case by Subgenic F1. In the case of Helixer, these scores represent cross-species predictions. We also compare the median performances of different configurations of Helixer with AUGUSTUS and a dilated CNN baseline in Table \ref{tab:results}. The results show a clear improvement over the AUGUSTUS both in median performance and spread. We also significantly outperform a baseline constructed of a dilated CNN architecture.
